--- 
title: "The Seven Wonders of the World – a text on physics and natural philosophy"
author: "PierGianLuca  Porta Mana"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a textbook-like collection of notes on various topics in physics and natural philosophy."
---

# Preface {-}

This is a collection of notes on various topics in physics, especially: classical mechanics & thermodynamics (that is, thermomechanics), general relativity, quantum theory, statistical mechanics.

Parts of these notes could be material for a textbook, although the notes as a whole cannot be considered one. My idea is not to present a physics course from the ground up; this would lead to repeating material that can already be found in good textbooks. Rather, I would like to replace the presentation and development of some topics, common to many textbooks, with different ones that I have not found in any textbook.

The reason is that most undergraduate textbooks on physics -- even those which appear in brand-new "latest editions" with flashy covers -- contain very outdated and sometimes even incorrect ideas and notions.

It is due time to update them.

Some examples:

1. Classical Newtonian mechanics can today be presented through notions and maths that make it extremely close to general relativity. Such an alternative presentation makes many notions much clearer and easier to understand for students, and at the same time reduces to a smooth transition the conceptual jump that students usually have to make when studying special and general relativity.

2. What many undergraduate physics textbooks call "thermodynamics" is usually just thermo*statics*, and moreover a thermostatics presented through obscure and unwieldy maths. Results that are only valid in static situations, say, the famous formula "$p = n R T/V$", are  presented as if they were also valid in non-static situations. Results that are only valid for elastic systems -- that is, systems whose states can all be connected by reversible processes -- are presented as if they were also valid for non-elastic systems. One important example is the statement that "entropy is defined up to an additive constant", which is far from being valid for all thermodynamic systems; in fact, it is false for many common everyday systems.


<!-- 
```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
 -->
