# [Probability inference]{.green} {#sec-probability}
{{< include macros.qmd >}}
{{< include macros_prob_inference.qmd >}}

In most engineering and data-science problems we don't know the truth or falsity of outcomes and hypotheses that interest us. But this doesn't mean that nothing can be said or done in such situations. Now we shall finally see how to draw *uncertain* inferences, that is, how to calculate the *probability* of something that interests us, given particular data, information, and assumptions.

So far we have used the term "probability" somewhat informally and intuitively. It is time to make it more precise and to emphasize some of its most important aspects, especially for Artificial Intelligence. Then we'll dive into the rules of probability-inference.


## When truth isn't known: beliefs and probability {#sec-probability-def}

If we want to design an AI agent that can make decisions and act in uncertain situations, we need to equip it with *probabilities* and *utilities*, as we saw in [ch. @sec-framework]. It is intuitively clear what utilities are: they quantify how desirable the agent finds different outcomes. What do probabilities quantify?

Think of situations in which you had to choose between several possible decisions, each having uncertain outcomes. What drove your decision was the desiderability or undesiderability of the outcomes, and also how strongly you *believed* in the occurrence of possible outcomes. For instance, you have to walk outdoors; it looks like it could rain; you consider whether to take an umbrella, or just a rain jacket, or just a normal jacket. You weigh the (un)desiderability of getting just a couple light raindrops on your head, getting soaked, or staying dry. You also assess your belief that it won't rain, or it will just shed a couple light drops, or it will rain heavily. Maybe you end up thinking "It looks like it won't rain yet in the next ten minutes, so I can just walk to my destination without the umbrella".

In building an AI agent that can act under uncertainty, we likewise need the agent to be able to [**quantify its beliefs**]{.blue} and its uncertainty about the possible outcomes. [*This is what probabilities are*]{.blue}.

:::{.callout-note}
##  

::::{style="font-size:120%"}

The [**probability**]{.blue} of a sentence is an agent's quantified [**degree of belief**]{.blue} in the truth of that sentence.

::::
:::

We shall take the notion of [*degree of belief*]{.blue} as intuitively understood, just as we did with the notion of truth. And we shall use the terms [*probability*]{.blue}, [*degree of belief*]{.blue}, [*plausibility*]{.blue}, [*credibility*]{.blue}^[*credibility* literally means "believability" (from Latin *credo* = *to believe*).] as synonyms.

\

Probabilities are quantified between $0$ and $1$, or equivalently between $0\%$ and $100\%$. Assigning to a sentence a probability `1` is the same as saying that it is `true`; and a probability `0`, that it is `false`. A probability of `0.5` represents a belief completely symmetric with respect to truth and falsity.

Said otherwise, if an agent assigns to a sentence a probability `1`, it means that the agent is completely certain that the sentence is `true`. If the agent assigns a probability `0`, it means that the agent is completely certain that the sentence is `false`. If the agent assigns a probability `0.5`, it means that the agent is equally uncertain about the truth as about the falsity of the sentence.

Let's emphasize and agree on some important facts about probabilities:

### [{{< fa hand-point-right >}} Probabilities are not frequencies]{.blue}

In thinking about the rainy-weather example above, you might say: "If I have a strong belief that it will rain, it's because in past similar situations, when I observed a similar sky and temperature condition, it ended up raining in the majority of those situations". Likewise, in our initial electronic-component scenario, you might say "those probabilities must be the frequencies with which previous electronic components failed or worked".

Indeed there is a connection between degree of belief and frequency. We shall *quantitatively* discuss this connection in later chapters, and make sure that an AI agent uses this connection.

Yet, degree of belief and frequency are not exactly the same, and in designing an AI agent it is important to keep the two notions separate. Let's see this with an example:
\

> Suppose someone shows you and a fried of yours a mechanical device designed to toss a coin. The device looks complicated and neither of you has a full grasp of its design. A normal coin is placed on the device, and your friend is required to place a $1 bet on whether the tossed coin will land `heads` or `tails`. You two cannot discuss, you can only observe your friend's choices. Note that not betting is not an option: feel free to imagine circumstances (possibly very nasty) where your friend can't refuse to bet; and consider that we're examining this example also from the point of view of an *AI* agent that might have to choose between the two options, with no other options available. Let's also make clear that no "cheating" is taking place: for instance, the outcome does *not* depend on the betting choice (you can imagine that the coin is tossed right before your friend bets, but the outcome hidden from them until they bet).
> 
> In this situation your friend says that it does not matter whether they bet on `heads` or `tails`; in fact, they decide by tossing a coin they had in their pocket.
> 
> *Do you think your friend's choice is *rational*?*
> 
> Their conclusion comes not only from the fact that the win is equal to the loss, but also from the fact that **their beliefs in the `heads` outcome and in the `tails` outcome are equal**. If one belief were stronger than the other, your friend would definitely bet on that outcome instead.
> 
> Yet this equality of beliefs *does not come from an equality of frequencies*. In fact neither you or your friend have any frequencies whatsoever. And for all you know, the device might be designed to always produce `heads`, or always `tails`, or maybe one or the other with peculiar frequencies not equal to 50%/50%.  
> 
> 
> Now suppose that after the first bet you and your friend observe the outcome, say `tails`. The device is thereafter repeatedly used, say for 1000 times (plus some extra times until you or your friend says "stop", if you like). In these repetitions you and your friend count that `heads` occurred 781 times, and `tails` 219 times, without any recognizable pattern. Now your friend is asked again to place a bet on the next toss (and remember: no one is trying to deceive you; no cheating).
> 
> Your friend says that they believe more strongly that the coin will land `heads` than `tails`. They even quantify their belief to around 78% for `heads` and 22% for `tails`. It's very close or identical to the frequency you both observed. Your friend bets accordingly.
> 
> *Do you think your friend's judgement and choice are *rational*?*
> 
> Finally, now a new friend of yours is brought in, and shown the device. This new friend cannot communicate with you two, and they are in the same situation you and your other friend were around 1000 tosses ago. The new friend is required to place a bet, and they say that they have no preferences for betting on `heads` or `tails` -- with exactly the same reasoning your other friend did 1000 tosses ago.

> *Do you think your new friend's choice is *rational*?*

\

This example brings to light that the beliefs underlying a rational choice may be different from frequencies, even though in some situations there can be a strong connection between the two. We therefore must keep these two notions quite distinct.

This distinction is especially important in the design of an AI agent. Just like us, an AI agent may have to make a decision, and therefore quantify the plausibility of different outcomes, when no frequencies are known, or more generally in situations that are unique, with no previous examples. We need to have a number to quantify the plausibility that the agent attributes to the possible outcomes. Such a number cannot be a frequency, because that may not be available. [*This number, the agent's degree of belief, is what we call "probability".*]{.blue}

But obviously we want such a number to be equal or close to a frequency, when it's available. This is not a problem: we shall see that the framework we're going to develop will do this automatically!
\

Note also the different status of these two notions from an epistemological point of view.^[That is, from the point of view of an agent's knowledge.] Frequencies can be unknown to some agents. Degrees of belief cannot be "unknown": the agent must have them in order to act. At worst, degrees of belief can be just difficult to calculate.

:::{.callout-important}
## {{< fa exclamation-triangle >}}

Be careful when you read authors speaking of an "unknown probability": they actually mean either "unknown frequency", or a probability that has to be calculated; it's "unknown" in the same sense that the value of\ \ "$1-0.7 \cdot 0.2/(1-0.3)$"\ \ is "unknown" to you right now.
:::

### [**{{< fa hand-point-right >}} Probabilities are agent- and knowledge-dependent**]{.blue}

The coin-tossing-device example above also shows that different agents can have different probabilities, that is, degrees of belief, about the same situation. 

This happened when your second friend entered the scene. You and your first friend had beliefs around 78% for `heads` and 22% for `tails`; your second friend had 50%/50% beliefs. Both sets of beliefs were rational and appropriate to their respective situations. They simply reflected the different states of knowledge of the agents that held them.

An omniscient agent would know the truth or falsity of every sentence, and assign only probabilities `0` or `1`. Some authors speak of "*actual* (but unknown) probabilities". But if there were "actual" probabilities, they would be all `0` or `1`, and it would be pointless to speak about probabilities at all -- every inference would be a truth-inference.


### [**{{< fa hand-point-right >}} Probabilities are not physical properties**]{.blue}

The fact that two agents can hold different probabilities in the same situation also shows that probabilities are not physical properties, which could me objectively measured with some meter.

Whether a tossed coin lands heads up or tails up is fully determined by the initial conditions (position, orientation, momentum, rotational momentum) of the toss and the boundary conditions (air velocity and pressure) during the flight. The same is true for all macroscopic engineering phenomena (even quantum phenomena have never been proved to be non-deterministic, and there are [deterministic and experimentally consistent](https://doi.org/10.48550/arXiv.quant-ph/9504010)  mathematical representations of quantum theory). So we cannot measure a probability using some physical apparatus.

::::{.column-margin}
::: {.callout-warning}
## {{< fa book >}} Study reading
[**Skim through**]{.lightblue} Diaconis & al. 2007: [*Dynamical Bias in the Coin Toss*](references.html).
<!-- We conclude that coin tossing is "physics" not "random." -->
:::
::::

We can objectively measure frequencies, in several instances of a phenomenon. Frequencies, as opposed to probabilities, are physically measurable quantities. This shows again the difference between probabilities and frequencies.



### [**{{< fa hand-point-right >}} Probabilities are assigned to *sentences***]{.blue}

We already discussed this point in [§@sec-sentence-notation], but let's reiterate it. Consider an engineer working on a problem of electric-power distribution in a specific geographical region. At a given moment the engineer may believe with $75\%$ probability that the measured average power output in the next hour will be 100 MW. The $75\%$ probability is assigned not to the quantity "100 MW", but to the *sentence*

$$
\pr{The measured average power output in the next hour will be 100\,MW}
$$

This difference is extremely important. Consider the alternative sentence

$$
\pr{The average power output in the next hour will be \emph{set} to 100\,MW}
$$

the numerical quantity is the same, but the meaning is very different. The probability can therefore be very different. If the engineer is the person who decides how to set that output, and has decided to set it to 100 MW, then the probability is obviously $100\%$ (or very close to), because the engineer already knows what the output will be. The probability depends not only on a number, but on what it's being done with that number: measuring, setting, third-party reporting, and so on. Often we write simply "$O \mo 10\,\mathrm{W}$", provided that the full sentence behind this shorthand is understood.

\

These points listed above are not just a matter of principle. They have important practical consequences. A data scientist who is not attentive to the source of the data (measured? set? reported, and so maybe less trustworthy?), or who does not carefully assess the context of a probability, or who mixes a probability with a frequency, or who does not take advantage (when possible) of the physics involved in the a problem -- such data scientist will design systems with sub-optimal performance^[This fact can be mathematically proven.] -- or even cause deaths.


## The many uses of the word "probability"

In these notes we shall consistently use the term "probability" in the sense explained above: an agent's *degree of belief*. But beware that this term is used in many different and incompatible senses, depending on whom you're speaking with or which literature you're reading.

Some people use this term in the sense of "frequency": the number of times something happened in a series of repetitions. As we have seen, to build an AI we need to quantify both notions and keep them separate: the distinction between "frequency" and "probability" is crucial. Frequencies do not enable us to quantify an agent's beliefs in situations where data are missing. But an agent's beliefs will automatically come close to frequencies when doing so is reasonable.
\
\

Some people use "probability" in the sense of the number of *a-priori* successes over the number of possibilities. For instance, if you roll a die, and the die comes up {{< fa dice-three >}}, then this result was 1 among six possible results. This is fine, but this definition does not allow us to capture the difference between an agent who has seen that the outcome of the roll was {{< fa dice-three >}}, and an agent who has *not* seen that outcome. We expect the two agents to behave differently, because they have different knowledge. If we programmed the agent to have a degree of belief of 1/6 *even after seeing the outcome of a die roll*, we would have programmed an agent incapable of using new data. Would you be happy if a clinician made some medical tests, and then *ignored* the results of the tests, behaving as if they were still unknown?
\
\

All these different uses are just a matter of semantics, and in the end it doesn't matter which word we use, as long as we understand its meaning and as long as we're adopting the meaning that is *useful* for our present task.

:::{.callout-important}
## {{< fa exclamation-triangle >}} Beware of *likelihood* as a synonym for *probability*

In everyday language, "[likelihood](https://dictionary.cambridge.org/dictionary/english/likelihood)" is synonym with "probability". In technical writings about probability or statistics, however, "likelihood" means something different and is *not* a synonym of "probability", as we explain below ([§@sec-likelihood]).
:::


## An unsure inference. Probability notation {#sec-uncertain-inference}

Consider now the following variation of the trivial inference problem of [§@sec-trivial-inference].

> This electric component had an early failure. If an electric component fails early, then at production either it didn't pass the heating test or it didn't pass the shock test. The probability that it passed neither test (that is, *both* tests failed) is 10%. There's no reason to believe that the component passed the heating test, more than to believe that it passed the shock test.

Again the inspector wants to assess whether the component *did not pass* the heating test.

From the data and information given, what would you say is the probability that the component didn't pass the heating test?

:::{.callout-caution}


- Try to argue why a conclusion cannot be drawn with certainty in this case. One way to argue this is by presenting two different scenarios that fit the given data but have opposite conclusions.

- Try to reason intuitively and assess the probability that the component didn't pass the heating test. Should it be larger or smaller than 50%? Why?
:::

For this inference problem we cannot find a `true` or `false` final value. The truth-inference rules (@eq-t-not)--(@eq-t-unity) therefore cannot help us here. In fact even the "$\mathrm{T}(\dotso \| \dotso)$" notation is unsuitable, because it only admits the values $1$ (`true`) and $0$ (`false`).

Let us first generalize this notation in a straightforward way:

First, let's represent the probability or degree of belief of a sentence by a number in the range [$[0,1]$,]{.m} that is, between $\mathbf{1}$ (certainty or `true`) and $\mathbf{0}$ (impossibility or `false`). The value $0.5$ represents a belief in the truth of the sentence which is as strong as the belief in its falsity.

Second, let's symbolically write in the following way that the probability of a proposal $\yY$, given a conditional $\yX$, is some number $p$:

$$
\P(\yY \| \yX) = p
$$

Note that this notation includes the notation for truth-values as a special case:

$$
\P(\yY \| \yX) = 0\text{ or }1
\quad\Longleftrightarrow\quad
\mathrm{T}(\yY \| \yX) = 0\text{ or }1
$$

## Inference rules {#sec-fundamental}

Extending our truth-inference notation to probability-inference notation has been straightforward. But which rules should we use for drawing inferences when probabilities are involved?

The amazing result is that *the rules for truth-inference, formulae (@eq-t-not)--(@eq-t-unity), extend also to probability-inference*. The only difference is that they now hold for all values in the range $[0,1]$, rather than only for $0$ and $1$.

This important result was taken more or less for granted at least since Laplace in the 1700s, but was formally proven for the first time in 1946 by R. T. Cox. The proof has been refined since then. What kind of proof is it? It shows that if we don't follow the rules we are doomed to arrive at illogical conclusions; we'll show some examples later.

\
Finally, here are *the fundamental rules of all inference*. They are encoded by the following equations, which must always hold for any atomic or composite sentences $\yX,\yY,\yZ$:

::::: {.column-page-right}
:::: {.callout-note style="text-align:center;font-size:125%"}
## [**{{< fa landmark >}}\ \ THE FUNDAMENTAL LAWS OF INFERENCE\ \ {{< fa landmark >}}**]{style="font-size:150%"}

:::{style="font-size:165%"}
$\boldsymbol{\lnot}$ "Not" rule
: $$\P(\lnot \yX \| \yZ) 
= 1 - \P(\yX \| \yZ)$$
\

$\boldsymbol{\land}$ "And" rule
: $$
\P(\yX \land \yY \| \yZ) 
= \P(\yX \| \yY \land \yZ) \cdot
\P(\yY \| \yZ) 
= \P(\yY \| \yX \land \yZ) \cdot
\P(\yX \| \yZ) 
$$
\

$\boldsymbol{\lor}$ "Or" rule
: $$\P(\yX \lor \yY \| \yZ) 
= \P(\yX \| \yZ) +
\P(\yY \| \yZ) 
- \P(\yX \land \yY \| \yZ)
$$
\

Truth rule
: $$\P(\yX \| \yX \land \yZ) 
= 1
$$
:::
\
\

::: {style="text-align:left;font-size:100%"}
**How to use the rules**:\
Each equality can be rewritten in different ways according to the usual rules of algebra. Then the resulting left side can be replaced by the right side, and vice versa. The numerical values of starting inferences can be replaced in the corresponding expressions.
:::
::::
:::::


It is amazing that **ALL** inference is nothing else but a repeated application of these four rules -- maybe billions of times or more. All machine-learning algorithms are just applications or approximations of these rules. Methods that you may have heard about in statistics are just specific applications of these rules. Truth inferences are also special applications of these rules. Most of this course is just a study of how to apply these rules to particular kinds of problems.

::: {.callout-warning}
## {{< fa book >}} Study reading

[**Read**]{.lightblue}:

- Ch. 2 of Gregory: [*Bayesian Logical Data Analysis for the Physical Sciences*](references.html)

- Ch. 1 of O'Hagan: [*Probability*](references.html)

- §§1.0--1.2 of Sivia: [*Data Analysis*](references.html)

[**Skim through**]{.lightblue}:

- Cox 1946: [*Probability, Frequency and Reasonable Expectation*](references.html). Try to get the ideas behind the reasoning, even if you can't follow the mathematical details.

- Chs 1--2 of Jaynes: [*Probability Theory*](references.html)
:::

\
The fundamental inference rules are used in the same way as their truth-inference counterpart of [§@truth-inference-rules]: Each equality can be rewritten in different ways according to the usual rules of algebra. The left and right side of the equality thus obtained can replace each other in a proof.

## Solution of the uncertain-inference example {#sec-solution-first-probinference}

Armed with the fundamental rules of inference, let's solve our earlier inference problem. As usual, we first analyse it and represent it in terms of atomic sentences; we find what are its proposal and conditional; and we find which initial inferences are given in the problem.

### 1. Atomic sentences

$$
\begin{aligned}
\yh &\defd \pr{The component passed the heating test}
\\
\ys &\defd \pr{The component passed the shock test}
\\
\yf &\defd \pr{The component had an early failure}
\\
\yJ &\defd \prq{(all other implicit background information)}
\end{aligned}
$$

The background information in this example is different from the previous, truth-inference one, so we use the different symbol $\yJ$ for it.

### 2. Proposal, conditional, and target inference

The proposal is $\lnot\yh$, just like in the [truth-inference example](#sec-trivial-inference-analysis).

The conditional is different now. We know that the component failed early, but we don't know whether it passed the shock test. Hence the conditional is
$\yf \land \yJ$.

The target inference is therefore

$$
\red\P(\lnot\yh \| \yf \land \yJ)
$$

### 3. Starting inferences

We are told that if an electric component fails early, then at production it didn't pass the heating test or the shock test (or neither). This is given as a sure fact. Let's write it as

$$
\green\P(\lnot\yh \lor \lnot\ys \| \yf \land \yJ) = 1
$$ {#eq-start-inf-a}

We are also told that there is a $10\%$ probability that both tests fail:

$$
\green\P(\lnot\yh \land \lnot\ys \| \yf \land \yJ) = 0.1
$$ {#eq-start-inf-b}

Finally the problem says that there's no reason to believe that the component didn't pass the heating test, more than it didn't pass the shock test. This can be written as follows:

$$
\green\P(\yh \| \yf \land \yJ) = \P(\ys \| \yf \land \yJ)
$$ {#eq-start-inf-c}

Note the interesting situation above: we are not given the numerical values of these two probabilities; we are only told that they are equal. This is an example of application of the *principle of indifference*, which we'll discuss more in detail later.

### Finding the target inference

Also in this case there is no unique way of applying the rules to reach our target inference, but *all paths will lead to the same result*. Let's try to proceed backwards:

::: {.column-page-inset-right}
$$
\begin{aligned}
&\red\P(\lnot\yh \| \yf \land \yJ)&&
\text{\small ∨-rule}
\\[1ex]
&\qquad= {\green\P(\lnot\ys \lor \lnot\yh \| \yf \land \yJ)}
+ {\green\P(\lnot\ys \land \lnot\yh \| \yf \land \yJ)}
- \P(\lnot\ys \| \yf \land \yJ)
&&
\text{\small starting inferences (8.1–2)}
\\[1ex]
&\qquad= {\green1}
+ {\green0.1}
- \P(\lnot\ys \| \yf \land \yJ)
&&
\text{\small ¬-rule}
\\[1ex]
&\qquad= 0.1 + \green\P(\ys \| \yf \land \yJ)
&&
\text{\small starting inference (8.3)}
\\[1ex]
&\qquad= 0.1 + \P(\yh \| \yf \land \yJ)
&&
\text{\small ¬-rule}
\\[1ex]
&\qquad= 0.1 + 1 -\red\P(\lnot\yh \| \yf \land \yJ)
&&
\end{aligned}
$$
:::

The target probability appears on the left and right side with opposite signs. We can solve for it:

$$
\begin{aligned}
2\,{\red\P(\lnot\yh \| \yf \land \yJ)} &= 0.1 + 1
\\[1ex]
{\red\P(\lnot\yh \| \yf \land \yJ)} &= 0.55
\end{aligned}
$$

So the probability that the component didn't pass the heating test is $55\%$.

:::{.callout-caution}


- Try to find an intuitive explanation of why the probability is 55%, slightly larger than 50%. If your intuition says this probability is wrong, then:

    + Check the proof of the inference for mistakes, or try to find a proof with a different path.
    + Examine your intuition critically and educate it.

- Check how the target probability $\P(\lnot\yh \| \yf \land \yJ)$ changes if we change the value of the probability $\P(\lnot\ys \land \lnot\yh \| \yf \land \yJ)$ from $0.1$.

    + What result do we obtain if $\P(\lnot\ys \land \lnot\yh \| \yf \land \yJ)=0$? Can it be intuitively explained?
    + What if $\P(\lnot\ys \land \lnot\yh \| \yf \land \yJ)=1$? Does the result make sense?
:::


## Use and implementation of the inference rules {#sec-algorithm-rules}

In the [step-wise solution above](#sec-solution-first-probinference) you noticed that the equations of the fundamental rules were not only used to calculate the probability on their left side, given those on their right side. For example, in the very first step, when we went from the probability

$$\P(\lnot\yh \| \yf \land \yJ)$$

to the sum

$$\P(\lnot\ys \lor \lnot\yh \| \yf \land \yJ)
+ \P(\lnot\ys \land \lnot\yh \| \yf \land \yJ)
- \P(\lnot\ys \| \yf \land \yJ) \,,
$$

we used the or-rule rewritten, by simple algebra, as follows:

$$
\P(\yX \| \yZ)
= \P(\yX {\red\lor} \yY \| \yZ)
- \P(\yY \| \yZ)
+ \P(\yX \land \yY \| \yZ) \,.
$$

The four rules in fact represent, first of all, [*constraints of logical consistency*]{.blue} (the precise technical term is [**coherence**]{.blue}) among probabilities. For instance, if we have probabilities

$$\begin{aligned}
\P(\yY\|\yX\land \yZ) &= 0.1
\\
\P(\yX\|\yZ) &= 0.7
\\
\P(\yX\land \yY\|\yZ) &= 0.2 \,,
\end{aligned}$$

then there must be an inconsistency the agent's degrees of belief, because these values violate the and-rule:

$$0.2 \ne 0.1 \cdot 0.7 \,.$$

When this happens, the inconsistency must be found and solved. (However, since probabilities are quantified by real numbers, it's possible and acceptable to have slight discrepancies within numerical round-off errors.)

The fundamental rules also imply more general constraints. For example we must *always* have

$$
\begin{gathered}
\P(\yX\land \yY \|\yZ) \le \min\set[\big]{\P(\yX\|\yZ),\  \P(\yY\|\yZ)}
\\
\P(\yX\lor \yY \|\yZ) \ge \max \set[\big]{\P(\yX\|\yZ),\  \P(\yY\|\yZ)}
\end{gathered}
$$

::: {.callout-caution icon=false}

- Try to prove the two constraints above from the four fundamental rules.

- Translate the two constraints above verbally. Do they make intuitive sense? (For example, the first constraint says, very roughly speaking, that our belief in *two* things can't be stronger than our belief in either one of the two things alone.)
:::

\

In following the step-wise solution you may have been asking yourself after some steps: "OK, why this peculiar step now, and not some other step?". This is a very intelligent question. You essentially noticed the *lack of an algorithm*. Compare this situation with the [solution of a basic decision problem](#sec-make-decision): there we had a clear sequence of steps to do: writing down some tables, do some element-wise multiplications and then some sums, and finally find the largest of a list of numbers.

Is there an algorithmic way of drawing inferences, that is, of calculating some target probabilities given some initial ones? If our inference problem involves a *finite* number of sentences, then the answer is *yes*, but with some caveats.

Given a set of initial probabilities, and a target probability whose value we want to find, there is an algorithm that yields **the minimum and the maximum possible values** of the target probability. Specifically it can yield the following results (note that some are particular instances of others):

::::{.column-margin}
::: {.callout-tip}
## {{< fa rocket >}} For the extra curious
The complete proof that inference problems can be solved this way seems to have been given first by Hailperin:

- (1965): [*Best Possible Inequalities for the Probability of a Logical Function of Events*](references.html),

- (1996): [*Sentential Probability Logic: Origins, Development, Current Status, and Technical Applications*](references.html),

although particular cases were explored already by Boole.

:::
::::

[{{< fa right-from-bracket >}}\ \ \ *No values*]{.blue}
: This means that the initial probabilities have inconsistent values; that is, they violate some of the four rules, as in the "$0.2 \ne 0.1 \cdot 0.7$" example above.
\
\

[{{< fa right-from-bracket >}}\ \ \ *$0$ and $1$*]{.blue}
: That is, the target probability is completely unspecified. This means that the initial probabilities are not sufficient to determine the target probability. For example, if we have $\P(\yX \| \yZ) = 0.2$, and $\yY$ is a sentence completely unrelated to $\yX$ (no atomic sentences in common), then $\P(\yY \| \yZ)$ could have any value.
\
\

[{{< fa right-from-bracket >}}\ \ \ *Values $v$, $V$ with $v < V$ and at least one different from 0 or 1*]{.blue}
: That is, the target probability cannot have any value whatsoever, but a its value is still unspecified. This means that the initial probabilities constrain the target probability somewhat, but do not determine it. The two constraints discussed above are an example of this: if we have $\P(\yX \| \yZ) = 0.2$ and $\P(\yY \| \yZ) = 0.3$, the we can say that $0 \le \P(\yX \land \yY\| \yZ) \le 0.2$, but its precise value is not determined.
\
\

[{{< fa right-from-bracket >}}\ \ \ *One value $p$*]{.blue}
: The target probability is completely determined by the initial ones. This occurred in our example inference about the electronic component.
\
\

This algorithm is not mathematically difficult, but it is somewhat involved. It boils down to:\ (1) writing the sentences in the initial and target probabilities in [*disjunctive normal form*](https://mathworld.wolfram.com/DisjunctiveNormalForm.html);\ (2) rewriting the initial and target probabilities as sum of probabilities of basic conjunctions;\ (3) solving two *linear-fractional optimization* problems, equivalent to [*linear optimization*](https://mathworld.wolfram.com/LinearProgramming.html) oens (for which there are algorithms available).

## An implementation in R

The algorithm described in the previous section is implemented in the R function [`inferP()`](https://github.com/pglpm/ADA511/blob/master/code/inferP.R); it requires the [**lpSolve**](https://cran.r-project.org/package=lpSolve) R package, which we installed in the [R introduction](Rintro1.html).

The `inferP()` function takes as first `target =` argument the probability we want to find, and as remaining arguments the values of the given probabilities, or equalities among probabilities. We must use the following notation:

- $\lnot$\ \ becomes\ \ ` -`
- $\land$\ \ becomes\ \ ` & `\ or equivalently\ ` * `
- $\lor$\ \ becomes\ \ `+`
- $=$\ \ becomes\ \ `==`
- conditional bar $\|$\ \ becomes\ \ ` | `

The output are the minimum and maximum possible values of the target probability, or `NA` if the starting probabilities are inconsistent.

We can apply this function in the example above about the electronic component:

```{r}
#| echo: false
knitr::opts_knit$set(root.dir = "other_code/")
```

```{r}
source('inferP.R') ## load the function

inferP(
    ## target probability:
    target = P(-h | f & J),
    ## given probabilities:
    P(-h + -s | f & J) == 1,
    P(-h & -s | f & J) == 0.1,
    P(h | f  & J) == P(s | f & J)
)
```

Obviously the algorithm becomes more expensive, the larger the number of sentences in the inference problem. In inference problems involving continuous quantities, such as energy, such an algorithm cannot be applied in practice. Also, the way the algorithm above works cannot be represented as a sequence of "logical" steps with consecutive applications of the rules, as shown in the step-wise solution above. Later on in this course we shall focus on specific kinds of inferences for which other, less opaque inference algorithms are available.

:::{.callout-caution}

- Play with the function `inferP()`: test it in other inference problems, find problems where there is no solution and others where the `min` and `max` values are different.

- Consider the following inference:
```{r}
inferP(
    target = P(hypothesis  |  evidence1 & evidence2 & I),
    P(hypothesis  |  evidence1 & I) == 0.1,
    P(hypothesis  |  evidence2 & I) == 0.9
)
```
What does this result tells us about "combining evidence" to prove a hypothesis?
:::

## Consequences of not following the rules

The fundamental rules of inference guarantee that the agent's uncertain reasoning is self-consistent, and that it follows logic when there's no uncertainty. Breaking the rules means that the resulting inference has some logical or irrational inconsistencies.

There are many examples of inconsistencies that appear when the rules are broken. Imagine for instance an agent that gives an 80% probability that it rains^[to be precise, let's say "it rains above 1 mm"] in the next hour; and it also gives a 90% probability that it rains *and* that the average wind is above 3⋅m/s in the next hour. This is clearly unreasonable, because the raining scenario alone would be true with wind above 3 m/s *and also* below 3⋅m/s -- therefore it should be *more* probable than the scenario where the wind is above 3 m/s. And indeed the two given probabilities break the `and`-rule, showing that they are unreasonable or illogical.

:::{.callout-caution}


Prove that the two probabilities in the example above break the `and`-rule.\ \ (Hint: you must use the fact that probabilities are numbers between 0 and 1, and that multiplying a number by something between 0 and 1 can only yield a smaller number.)
:::

::: {.callout-warning}
## {{< fa book >}} Study reading

- [**Read**]{.lightblue} §12.2.3 in [*Artificial Intelligence*](references.html)

- As you continue your studies, [**skim through**]{.lightblue} chapters 4--8 of Hastie & Dawes: [*Rational Choice in an Uncertain World*](references.html), just to get the main messages and an overview of curious psychological phenomena.
:::



<!-- @@ §12.2.3 of AI -->

<!-- * _Exercise: [Monty-Hall problem & variations](The_Monty_Hall_problem-exercise.pdf)_ -->

<!-- * _Exercise: clinical test & diagnosis_ -->

## Remarks on terminology and notation

### Likelihood {#sec-likelihood}

In everyday language, "likely" is often a synonym of "probable", and "likelihood" of "probability". But in technical writings about probability, inference, and decision-making, "likelihood" has a very different meaning. Beware of this important difference in definitions:

$\P(\yY\|\yX)$ is:

- the [**probability of $\yY$ given $\yX$**]{.blue} (or **conditional on $\yX$**),

- the [**likelihood of $\yX$ in view of $\yY$**]{.blue}.

We can also say:

- the [**probability of $\yY$**]{.blue} given $\yX$, is $\P({\blue\yY}\|\yX)$.

- the [**likelihood of $\yY$**]{.blue} in view of $\yX$, is $\P(\yX\|{\blue\yY})$.


:::{.callout-important}
## {{< fa exclamation-triangle >}} Probability and likelihood can be very different

A priori there is no relation between the probability and the likelihood of a sentence $\yY$: this sentence could have very high probability and very low likelihood, and vice versa.
:::

In these notes we'll avoid the possibly confusing term "likelihood". All we need to express can be phrased in terms of probability.

### Omitting background information

In the analyses of the inference examples of [§@sec-trivial-inference] and [§@sec-uncertain-inference] we defined sentences ($\yI$ and $\yJ$) expressing all background information, and always included these sentences in the conditionals of the inferences -- because those inferences obviously depended on that specific background information.

In many concrete inference problems the background information usually stays in the conditional from beginning to end, while the other sentences jump around between conditional and proposal as we apply the rules of inference. For this reason the background information is often omitted from the notation, being implicitly understood. For instance, if the background information is denoted $\yI$, one writes

- "$\P(\yY \|\yX)$"\ \ instead of\ \ $\P(\yY \|\yX \land \yI)$

- "$\P(\yY)$"\ \ instead of\ \ $\P(\yY \|\yI)$

This is what's happening in books where you see "$P(x)$" without conditional.

Such practice may be convenient, but be wary of it, especially in particular situations:

- In some inference problems we suddenly realize that we must distinguish between cases that depend on hypotheses, say $\yH_1$ and $\yH_2$, that were buried in the background information $\yI$. If the background information $\yI$ is explicitly reported in the notation, this is no problem: we can rewrite it as

    $$ \yI = (\yH_1 \lor \yH_2) \land \yI'$$

    and then proceed as usual. If the background information was not explicitly written, this may lead to confusion and mistakes: there may suddenly appear two instances of $\P(\yX)$ with *different* values, just because one of them is invisibly conditional on $\yI$, the other on $\yI'$.

- In some inference problems we are considering *several different* instances of background information -- for example because more than one agent is involved. It's then extremely important to write the background information explicitly, lest we mix up the degrees of belief of different agents.

::::{.column-margin}
::: {.callout-tip}
## {{< fa rocket >}} For the extra curious
A once-famous paper in the quantum-theory literature: Brukner & Zeilinger 2000: [Conceptual inadequacy of the Shannon information in quantum measurements](references.html) arrived at completely wrong results simply by omitting background information, mixing up probabilities having different conditionals.
:::
::::

This kind of confusion from poor notation happens more often than one thinks, and even appears in the scientific literature.


### "Random variables"

Some texts speak of the probability of a "random variable", or more precisely of the probability "that a random variable takes on a particular value". As you notice, we have just expressed that idea by means of a *sentence*. The viewpoint and terminology of random variables is therefore a special case of that based on sentences, which we use here.

The dialect of "random variables" does not offer any advantages in concepts, notation, terminology, or calculations, and it has several shortcomings:

::::{.column-margin}
![[James Clerk Maxwell](https://clerkmaxwellfoundation.org/html/about_maxwell.html) is one of the main founders of statistical mechanics and kinetic theory (and electromagnetism). Yet he never used the word "random" in his technical writings. Maxwell is known for being very clear and meticulous with explanations and terminology.](maxwell1.jpg){width=100%}
::::

- As discussed in [§@sec-probability-def], in concrete applications it is important to know how a quantity "takes on" a value: for example it could be directly measured, indirectly reported, or purposely set to that specific value. Thinking and working in terms of sentences, rather than of random variables, allows us to account for these important differences.

- We want a general AI agent to be able to deal with uncertainty and probability also in situations that do not involve mathematical sets.

- Very often the object (proposal) of a probability is not a "variable": it is actually a *constant* value that is simply unknown (simple example: we are uncertain about the mass of a particular block of concrete, so we speak of the probability of some mass value; this doesn't mean that the mass of the block of concrete is changing).

- What does "random" (or "chance") mean? Good luck finding an understandable and non-circular definition in texts that use that word. Strangely enough, texts that use that word never define it. In these notes, if the word "random" is ever used, it stands for "unpredictable" or "unsystematic".

It's a question for sociology of science why some people keep on using less flexible points of view or terminologies. Probably they just memorize them as students and then a fossilization process sets in.

\
Finally, some texts speak of the probability of an "event". For all purposes, an "event" is just what's expressed in a sentence.

